---
title: "DivvyAnalaysis"
author: "Sajith"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

# STEP 1: Importing libraries

```{r}
library(lubridate)
library(tidyverse)
library(ggplot2)
library("anytime")
```

# STEP 2: Collecting Data

```{r}
q2_2019 <- read.csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read.csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read.csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read.csv("Divvy_Trips_2020_Q1.csv")
```

# STEP 3: Wrangle Data AND COMBINE INTO A SINGLE FILE
 
### Comparing cols of each file
### the names are not in same order so we change them into same col heads throughout
### then we combine all of these files into a single new csv file

```{r}
colnames(q1_2020)
colnames(q4_2019)
colnames(q3_2019)
colnames(q2_2019)
```

# STEP 4: Renaming,Mutating, and Transformation of data

### Renaming other files using the same col heads as this  q1_2020 dataset

```{r include=FALSE}
(q4_2019 <- rename(q4_2019
                  ,ride_id=trip_id
                  ,rideable_type=bikeid
                  ,started_at=start_time
                  ,ended_at=end_time
                  ,start_station_name=from_station_name
                  ,start_station_id=from_station_id
                  ,end_station_name=to_station_name
                  ,end_station_id=to_station_id
                  ,member_casual=usertype))


(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))


(q2_2019 <- rename(q2_2019
                   ,ride_id = "X01...Rental.Details.Rental.ID"
                   ,rideable_type = "X01...Rental.Details.Bike.ID" 
                   ,started_at = "X01...Rental.Details.Local.Start.Time"  
                   ,ended_at = "X01...Rental.Details.Local.End.Time"  
                   ,start_station_name = "X03...Rental.Start.Station.Name" 
                   ,start_station_id = "X03...Rental.Start.Station.ID"
                   ,end_station_name = "X02...Rental.End.Station.Name" 
                   ,end_station_id = "X02...Rental.End.Station.ID"
                   ,member_casual = "User.Type"))
```

### Inspecting the dataframes for incongruencies

```{r}
str(q1_2020)
str(q4_2019)
str(q3_2019)
str(q2_2019)
```

### Convert ride_id and rideable_type to character so that they can stack correctly

```{r}
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
```

### Stacking individual dataframes into one big dataframe

```{r}
all_trips <- bind_rows(q1_2020,q2_2019,q3_2019,q4_2019)
```

### Remove lat, long, birthyear, and gender fields as this data was dropped beginning in 2020

```{r}
all_trips <- all_trips %>% 
  select(-c(start_lat,start_lng,end_lat,end_lng,gender,"X01...Rental.Details.Duration.In.Seconds.Uncapped",
            Member.Gender,tripduration,"X05...Member.Details.Member.Birthday.Year"))

all_trips <- all_trips %>% 
  select(-c(birthyear))
```

# STEP 5: CLEANING UP DATA AND AND ADD DATA TO PREPARE FOR ANALYSIS

###  Inspecting new data fram created for further analysis

```{r}
nrow(all_trips) #How many rows are in data frame?

colnames(all_trips) #List of column names

dim(all_trips)#Dimensions of the data frame?

head(all_trips)#See the first 6 rows of data frame.

tail(all_trips)#See the last 6 rows of data frame.

str(all_trips)#See list of columns and data types (numeric, character, etc)

summary(all_trips)
```

### There are a few problems we will need to fix:
      
#### (1) In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.
          
#### (2) The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.
          
#### (3) We will want to add a calculated field for length of ride since the  2020 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.
          
#### (4) There are some rides where tripduration shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides.


#### In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual"

#### Before 2020, Divvy used different labels for these two types of riders ... we will want to make our dataframe consistent with this format


```{r}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))
```

#### checking if it changed

```{r}
table(all_trips$member_casual)
```

#### Add columns that list the date, month, day, and year of each ride

#### This will allow us to aggregate ride data for each month, day, or year ... before completing

```{r}
all_trips$date <- as.Date(all_trips$started_at)

all_trips$month <- format(as.Date(all_trips$date), "%m")

all_trips$day <- format(as.Date(all_trips$date), "%d")

all_trips$year <- format(as.Date(all_trips$date), "%Y")

all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

#### Add a "ride_length" calculation to all_trips (in seconds)

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

#### Inspecting the structure of the columns

```{r}
str(all_trips)
head(all_trips)

```

#### Convert "ride_length" from Factor to numeric so we can run calculations on the data

```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(all_trips$ride_length)
is.numeric(all_trips$ride_length)

str(all_trips)
```

#### Removing Bad Data

#### The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative

#### We will create a new cleaned dataframe

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name=="HQ QR"|all_trips$ride_length<0),]
```

# STEP 6: CONDUCT DESCRIPTIVE ANALYSIS

#### Descriptive analysis on ride_length (all figures in seconds)

```{r}
mean(all_trips$ride_length) #straight average (total ride length / rides)

median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths

max(all_trips_v2$ride_length) #longest ride

min(all_trips_v2$ride_length) #shortest ride

summary(all_trips_v2$ride_length)
```

#### Compare members and casual users

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

#### See the average ride time by each day for members vs casual users

```{r}
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual+all_trips_v2$day_of_week, FUN = mean)
```

#### Notice that the days of the week are out of order. Let's fix that.

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

```

#### Now, let's run the average ride time by each day for members vs casual users

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

#### analyze ridership data by type and weekday

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%                  #groups by usertype and weekday
  summarise(number_of_rides = n()                       #calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>%  # calculates the average duration
  arrange(member_casual, weekday)                       # sorts
```

# STEP 7: VISUALIZATION

####  Let's visualize the number of rides by rider type

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

#### Let's create a visualization for average duration

```{r}
all_trips_v2 %>% 
  mutate(weekday=wday(started_at,label = TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarise(number_of_rides=n(),
            average_duration=mean(ride_length)) %>% 
  arrange(member_casual,weekday) %>% 
  ggplot(aes(x=weekday,y=average_duration,fill=member_casual)) + 
  geom_col(position = "dodge")
```


























